---
title: "Reproducible Research Assignment 1"
subtitle: "Activity Monitoring"
author: "Corey Bort"
date: "Sunday, July 20, 2014"
output: html_document
---

In this document I will walk you through the process of preparing some analysis
on the [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
set provided.

## Loading and preprocessing the data
After downloading and unzipping the file we need to set R's working 
directory to the correct location. You will need to modify the directory
for your system as it will not match mine.
```{r}
setwd("D:/Users/cbort/Google Drive/Edu/Coursera/Reproducible Research/RepData_PeerAssessment1")
```
  
Now that the directory is set we can read in the contents of the csv file.
```{r}
df <- read.csv("activity.csv", sep=",", stringsAsFactors=FALSE)
```
  
Since the dates were imported as character strings we will need to change them 
into dates.
```{r}
df[,2] <- as.Date(strptime(df[,2],"%Y-%m-%d"))
```
  
## What is mean total number of steps taken per day?
In order to find the mean total number of steps per day we can use the following 
R code.
```{r}
per_day <- aggregate(steps~date, df, sum)
```
This will allow us to easily plot a histogram of steps per day.
```{r}
hist(per_day$steps, col="darkcyan", xlab="Steps Per Day", main="Histogram of Steps Per Day")
```
Additionally we are able to easily calculate the mean and median steps per day
are calculated as follows:
```{r}
mean_steps<-mean(per_day$steps)
median_steps<-median(per_day$steps)
```

## What is the average daily activity pattern?
To calculate the daily activity pattern we will need to modify the data to group 
by the interval and average the steps taken accross days. This can be done using
the R code:
```{r}
per_interval <- aggregate(steps~interval, df, mean)
```
This can then be used to plot the average steps taken throughout the day
```{r}
plot(per_interval$interval, per_interval$steps, type='l',
     col="darkcyan",
     main="Average Daily Activity",
     xlab="Daily Interval",
     ylab="Steps")
```
...and to determine that the interval with the most steps taken:
```{r}
per_interval$interval[which.max(per_interval$steps)]

```
  
## Imputing missing values
This data set contains a number of rows with NA values. To clean up the dataset
and determine the NA impact we can first calculate the number of rows with NA
```{r}
dim(df[!complete.cases(df),])[1]
```

We can then clean up the NAs by replacing them with the median time for that 
interval. To do this we should first make sure only the steps column has NA 
values
```{r}
sum(is.na(df$interval))
sum(is.na(df$date))
sum(is.na(df$steps))
```
This shows us that indeed only the "steps" field contains NA values.
So we can now eliminate them by replacing NA with the median time for that 
interval.
```{r}
cleaned_df <- df
median_per_interval <- aggregate(steps~interval, df, median)
for( i in 1:dim(df)[1]){
    if (is.na(df$steps[i])){
        interval <- cleaned_df[i,3]
        steps <- median_per_interval[median_per_interval$interval==interval,2]
        cleaned_df[i,1] <- steps[1]
    }
}
```

Now we will repeat our previous analysis to find:
```{r}
cleaned_per_day <- aggregate(steps~date, cleaned_df, sum)
hist(cleaned_per_day$steps, col="darkcyan", xlab="Steps Per Day", main="Histogram of Steps Per Day(Clean)")
clean_mean_steps <- mean(cleaned_per_day$steps)
clean_median_steps <- median(cleaned_per_day$steps)
```
And we can then compare this to the previous values to see:
```{r}
(mean_steps - clean_mean_steps) / mean_steps * 100
( median_steps - clean_median_steps) / median_steps * 100
```
and we can see this causese the mean to increase by ~10% and the median by 3.5%


## Are there differences in activity patterns between weekdays and weekends?
We will now add a column stating whether these are weekdays or weekends and plot it.
```{r}
cleaned_df$weekday <- ifelse(weekdays(cleaned_df$date) %in% c("Saturday","Sunday"),"weekend","weekday")
weekend_per_interval <- aggregate(steps~interval+weekday, cleaned_df, mean)
weekend <- weekend_per_interval[weekend_per_interval$weekday=="weekend",c(1,3)]
weekday <- weekend_per_interval[weekend_per_interval$weekday=="weekday",c(1,3)]
par(mfrow=c(2,1))
plot(weekend$interval, weekend$steps, type='l',
     col="darkcyan",
     main="Weekend",
     xlab="Daily Interval",
     ylab="Steps")
plot(weekday$interval, weekend$steps, type='l',
     col="red",
     main="Weekday",
     xlab="Daily Interval",
     ylab="Steps")
```
